Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections

### 2023-05-16 jss/cheem_paper.Rmd
small update in the abstract

---
documentclass: jss
author:
  - name: Nicholas Spyrison
    orcid: 0000-0002-8417-0212
    affiliation: Monash University
    # use this syntax to add text on several lines
    address: |
      | Dept of Human Centred Computing
      | Faculty of Information Technology
    email: \email{spyrison@gmail.com}
    url: https://nspyrison.netlify.app
  - name: Dianne Cook
    orcid: 0000-0002-3813-7155
    address: |
      | Dept of Econometrics and Business Statistics
    affiliation: Monash University
  - name: Przemyslaw Biecek
    orcid: 0000-0001-8423-1823
    affiliation: |
      | Warsaw University of 
      | Technology and University
title:
  formatted: "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  plain:     "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  # For running headers, if needed
  short:     "Exploring Local Explanations"
abstract: >
    The increased predictive power of nonlinear models comes at the cost of interpretability of its terms. This trade-off has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models use predictors to arrive at a prediction.  "Local explanations", which provide a point estimate of the linear variable importance in the vicinity of one observation, are one XAI method. These can be considered to be linear projections and can be further explored to understand the interactions between variables used to make predictions across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any observation and illustrate with examples from categorical (penguin species, chocolate types) and quantitative (soccer/football salaries, house prices) response models. The methods are implemented in the R package cheem, available on CRAN.
keywords:
  # at least one keyword must be supplied
  formatted: [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
  plain:     [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
bibliography: paper.bib
---

>>>>

---
documentclass: jss
author:
  - name: Nicholas Spyrison
    orcid: 0000-0002-8417-0212
    affiliation: Monash University
    # use this syntax to add text on several lines
    address: |
      | Dept of Human Centred Computing
      | Faculty of Information Technology
    email: \email{spyrison@gmail.com}
    url: https://nspyrison.netlify.app
  - name: Dianne Cook
    orcid: 0000-0002-3813-7155
    address: |
      | Dept of Econometrics and Business Statistics
    affiliation: Monash University
  - name: Przemyslaw Biecek
    orcid: 0000-0001-8423-1823
    affiliation: |
      | Warsaw University of 
      | Technology and University
title:
  formatted: "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  plain:     "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  # For running headers, if needed
  short:     "Exploring Local Explanations"
abstract: >
    The increased predictive power of nonlinear models comes at the cost of interpretability of its terms. This trade-off has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models use predictors to arrive at a prediction.  "Local variable attributions", which provide a point estimate of the linear variable importance in the vicinity of one observation, are one local XAI method. These can be considered to be linear projections and can be further explored to understand the interactions between variables used to make predictions across the predictive model surface as well as understand the structure of local explanations for observations, whether outlier are present, whether segments can be identified. Here we describe interactive linear interpolation used for exploration at any observation and illustrate with examples from categorical (penguin species, chocolate types) and quantitative (soccer/football salaries, house prices) response models. The methods are implemented in the R package cheem, available on CRAN.
keywords:
  # at least one keyword must be supplied
  formatted: [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
  plain:     [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
bibliography: paper.bib
---

### 2023-05-16 jss/cheem_paper.tex
small update in the abstract

\Abstract{
The increased predictive power of nonlinear models comes at the cost of
interpretability of its terms. This trade-off has led to the emergence
of eXplainable AI (XAI). XAI attempts to shed light on how models use
predictors to arrive at a prediction. ``Local explanations'', which
provide a point estimate of the linear variable importance in the
vicinity of one observation, are one XAI method. These can be considered
to be linear projections and can be further explored to understand the
interactions between variables used to make predictions across the
predictive model surface. Here we describe interactive linear
interpolation used for exploration at any observation and illustrate
with examples from categorical (penguin species, chocolate types) and
quantitative (soccer/football salaries, house prices) response models.
The methods are implemented in the R package cheem, available on CRAN.
}

@@

There is a clear need to extend the interpretability of black box
models. With techniques such as SHAP, LIME, Break-down, one can
calculate local explanations, i.e.~for every observation how strongly
variables affect the model's predictions. Surprisingly few techniques
allow us to understand the global distribution of these local
explanations. Unsupervised data exploration techniques applied to data
show how useful they are for identifying outliers, identifying segments
of observations or discovering correlations between variables. All of
these tasks can be performed for a set of explanations.

This paper provides a technique that builds on local explanations to
explore the variable importance local to an observation. The local
explanations of an attribution projection from which variable
contributions are varied using a radial tour. Several diagnostic plots
are provided to assist with understanding the sensitivity of the
prediction to particular variables. A global view shows the data space,
explanation space, and residual plot. The user can interactively select
observations to compare, contrast, and study further. Then the radial
tour is used to explore the variable sensitivity identified by the
attribution projection.

>>>>

\Abstract{
The increased predictive power of nonlinear models comes at the cost of
interpretability of its terms. This trade-off has led to the emergence
of eXplainable AI (XAI). XAI attempts to shed light on how models use
predictors to arrive at a prediction. ``Local variable attributions'',
which provide a point estimate of the linear variable importance in the
vicinity of one observation, are one local XAI method. These can be
considered to be linear projections and can be further explored to
understand the interactions between variables used to make predictions
across the predictive model surface as well as understand the structure
of local explanations for observations, whether outlier are present,
whether segments can be identified. Here we describe interactive linear
interpolation used for exploration at any observation and illustrate
with examples from categorical (penguin species, chocolate types) and
quantitative (soccer/football salaries, house prices) response models.
The methods are implemented in the R package cheem, available on CRAN.
}

@@

There is a clear need to extend the interpretability of black box
models. With techniques such as SHAP, LIME, Break-down, one can
calculate local explanations, i.e.~for every observation in the data.
These techniques quantify for each observation how strongly particular
variables affect the model's predictions. Surprisingly few techniques
allow us to understand the global distribution of these local
explanations. Unsupervised data exploration techniques applied to data
show how useful they are for identifying outliers, identifying segments
of observations or discovering correlations between variables. All of
these tasks can be performed for a set of explanations.

To address this challenge in this paper provides a technique that builds
on local explanations to explore the variable importance local to an
observation. The local explanations of an attribution projection from
which variable contributions are varied using a radial tour. Several
diagnostic plots are provided to assist with understanding the
sensitivity of the prediction to particular variables. A global view
shows the data space, explanation space, and residual plot. The user can
interactively select observations to compare, contrast, and study
further. Then the radial tour is used to explore the variable
sensitivity identified by the attribution projection.
